NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-05 Manuscript/manuscriptPS.tex

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup and glycerol backbone from experiments
    and simulations with different force fields.
    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data is without additional ions.
    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for some simulation values
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
the best performing models,
Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
Slipids reproduces also the feature that the $\beta$-carbon order parameter
in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

>>>>

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (dots).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    All the data for DOPS at 303~K and all the data for POPS at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for all simulation values except MacRog~K$^+$
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
the best performing models,
Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
Slipids reproduces also the feature that the $\beta$-carbon order parameter
in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which focuses on the PS headgroup.
